---
title: "DSfinalproject"
author: "Brenda Nava"
date: "2026-04-09"
output: html_document
---



```{r setup, message=FALSE, warning=FALSE}
library(dplyr)        # Data wrangling
library(readr)        # Reading semicolon-separated CSV files
library(ggplot2)      # Visualization
library(skimr)        # Clean variable summaries
library(corrplot)     # Correlation heatmap
library(nnet)         # Multinomial logistic regression
library(randomForest) # Random forest classification
library(tidyr)        # Data tidying
```


# Data Preparation: The dataset contains chemical measurements for red and white wines, including acidity, sugar, sulfur dioxide, density, pH, sulphates, alcohol, and a quality score. After combining both datasets, the final dataset contains 6497 rows and 15 columns (I did add the quality_100 column).

```{r}

# semicolon separator
red <- read_delim("data/winequality-red (1).csv", delim= ";")
white <- read_delim("data/winequality-white (1).csv", delim= ";")

# wine type
red$wine_type <- "red"
white$wine_type <- "white"

#combine datasets
wine <- bind_rows(red, white)

#convert quality to 100 point scale
wine <- wine %>%
  mutate(
    quality_factor= as.factor(quality),
    quality_100 = 50 + (quality * 5)
  )

# Summary of variables
summary(wine)
```
```{r}

head(wine)

```


# How do the chemical variables correlate with one another?

```{r}
numeric_var <- wine %>% select(where(is.numeric))
cor_matrix <- cor(numeric_var)


corrplot(cor_matrix, method = "color", type="upper", tl.col = "black", tl.cex = .7)
```


# Does converting wine quality to a 100‑point scale make the distribution more intuitive?
```{r}
ggplot(wine, aes(x=quality_100, fill= wine_type))+
  geom_histogram(binwidth = 5, color="white")+
  theme_minimal()+
  labs(title = "Distribution of wine quality 100 point-scale")

```
```{r}

```




